pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-31 manuscript/greenwell.tex
minor edits

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}. The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}. The result is displayed in Figure~\ref{fig:ice_boston}. \strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in this example, there is one curve for each observation in \code{boston}.
\begin{example}
# Use partial to obtain ICE curves
pred.ice <- function(object, newdata) predict(object, newdata)
age.ice <- partial(boston.rf, pred.var = "age", pred.fun = pred.ice)

@@

For illustration, we return to the Boston housing data. The code chunk below uses \pkg{caret} to train an XGBoost model to the \code{boston} data frame from Section~\ref{sec:boston} over a grid of tuning values (see \code{?xgboost::xgboost} for an explanation of the various tuning parameters). The optimal combination of tuning parameter values is selected via 5-fold cross-validation.
\begin{example}
library(caret)

# Setup for 5-fold cross-validation
ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)

# Grid of tuning values for XGBoost.
xgb.grid <- expand.grid(nrounds = c(1000, 2000, 3000),
                        max_depth = 1:4,
                        eta = c(0.001, 0.01, 0.1),
                        gamma = 0,
                        colsample_bytree = 1,
                        min_child_weight = 1,
                        subsample = c(0.5, 0.75, 1))

# Train an XGBoost model
set.seed(202)
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv,
                    method = "xgbTree",
                    metric = "RMSE",
                    trControl = ctrl,
                    tuneGrid = xgb.grid)
\end{example}
The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 3000}, \code{max\_depth = 3}, \code{eta = 0.01}, and \code{subsample = 0.75}; the final model achieved a cross-validated RMSE and $R^2$ of 2.789 and 0.906, respectively. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
# Display plots side by side (Figure 11)
grid.arrange(
  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
  ncol = 3
)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using XGBoost. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}

%Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}

>>>>

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}. The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}. The result is displayed in Figure~\ref{fig:ice_boston}. \strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in the following code chunk, there will be one curve for each observation in \code{boston}.
\begin{example}
# Use partial to obtain ICE curves
pred.ice <- function(object, newdata) predict(object, newdata)
age.ice <- partial(boston.rf, pred.var = "age", pred.fun = pred.ice)

@@

For illustration, we return to the Boston housing example. The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation. (After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
\begin{example}
# Tune an XGBoost model using 10-fold cross-validation
library(caret)  # functions related to classification and regression training
set.seed(202)  # for reproducibility
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
                    trControl = trainControl(method = "cv", number = 10),
                    tuneLength = 10)
\end{example}
The optimal model had a cross-validated $R^2$ of $0.902$ which was achieved using \code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
# Display plots side by side (Figure 11)
grid.arrange(
  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
  ncol = 3
)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}

Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}. However, this is not the case when using the \pkg{xgboost} package directly. For example, the following code fits the same model using the \code{xgboost} function.
\begin{example}
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
                      colsample_bytree = 0.8, min_child_weight = 1,
                      subsample = 0.9444444)
\end{example}
To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}
